Supplementary material

### 2022-04-08 submission/manuscript.rmd
update manuscript and supplement

---
title: |
 Individualized treatment effect was predicted best by modeling 
 baseline risk in interaction with treatment assignment
abstract: |
 **Objective**: To compare different risk-based methods for optimal prediction
 of individualized treatment effects. **Study Design and Setting**: We simulated
 RCT data using diverse assumptions for the average treatment effect, a baseline
 prognostic index of risk (PI), the shape of its interaction with treatment
 (none, linear, quadratic or non-monotonic), and the magnitude of
 treatment-related harms (none or constant independent of the PI). We predicted
 absolute benefit using: models with a constant relative treatment effect;
 stratification in quarters of the PI; models including a linear interaction of
 treatment with the PI; models including an interaction of treatment with a
 restricted cubic spline (RCS) transformation of the PI; an adaptive approach
 using Akaike’s Information Criterion. We evaluated predictive performance using
 root mean squared error and measures of discrimination and calibration for
 benefit. **Results**: The linear-interaction model and the RCS-interaction
 displayed robust performance across many simulation scenarios. The RCS-model
 was optimal when quadratic or non-monotonic deviations from a constant
 treatment effect were stronger, and when sample size was larger. The adaptive
 approach required larger sample sizes. Illustrations in the GUSTO-I trial
 confirmed these findings. **Conclusion**: An interaction between baseline risk
 and treatment assignment should be considered to improve treatment effect
 predictions.
author:
  - name: Alexandros Rekkas
    affiliation: 1
    correspondingauthor: true
  - name: Peter R. Rijnbeek
    affiliation: 1
  - name: David M. Kent
    affiliation: 2
  - name: Ewout W. Steyerberg
    affiliation: 3
  - name: David van Klaveren
    affiliation: 4
address:
  - code: 1
    address: Department of Medical Informatics, Erasmus Medical Center, Rotterdam, The Netherlands
  - code: 2
    address: Predictive Analytics and Comparative Effectiveness Center, Institute for Clinical Research and Health Policy Studies, Tufts Medical Center, Boston, Massachusetts, USA
  - code: 3
    address: Department of Biomedical Data Sciences, Leiden University Medical Center, Leiden, The Netherlands
  - code: 4
    address: Department of Public Health, Erasmus Medical Center, Rotterdam, The Netherlands
numbersections: true
linenumbers: false
journal: arXiv.org
keywords:
  - treatment effect heterogeneity
  - absolute benefit
  - prediction models
output: 
  rticles::elsevier_article:
    keep_tex: true
  bookdown::word_document2:
    reference_docx: reference.docx
geometry: margin=1.0in
date: false
toc: false
font-size: 11pt
header-includes:
  - \renewcommand*\familydefault{\sfdefault}
  - \usepackage{setspace}
  - \usepackage{amsmath}
  - \doublespacing
  - \usepackage{amssymb}
  - \usepackage{bm}
  - \usepackage{caption}
  - \usepackage{booktabs}
  - \date{}
  - \newcommand\given[1][]{\:#1\vert\:}
editor_options: 
  chunk_output_type: console
bibliography: references.bib
csl: jce.csl
---

@@

@@

Outcomes in the treatment arm were generated using 3 base scenarios: absent
treatment effect (OR = 1), moderate treatment effect (OR = 0.8) and strong
treatment effect (OR = 0.5). We started with simulating outcomes based on true
constant relative treatment effects for the 3 base scenarios. We then simulated
linear, quadratic and non-monotonic deviations from constant treatment effects
using:
$$lp_1 = \gamma_2(PI-c)^2 + \gamma_1(PI-c) + \gamma_0, $$
where $lp_1$ is the true linear predictor in the treatment arm, so that
$P(y=1|X, t_x=1) = \text{expit}(lp_1)$. Finally, we simulated scenarios where a
constant absolute harm is applied across all treated patients. In this case we
have $P(y=1|X,t_x=1) = \text{expit}(lp_1) + \text{harm}$.

@@

Second, we considered a model which assumes *constant relative treatment effect*
(constant odds ratio). Hence, absolute benefit is predicted from
$\hat{\tau}(\bm{x}) = \text{expit}(PI +\log(\text{OR}))$.

>>>>

---
title: |
 Individualized treatment effect was predicted best by modeling 
 baseline risk in interaction with treatment assignment
abstract: |
 **Objective**: To compare different risk-based methods for optimal prediction
 of individualized treatment effects. **Study Design and Setting**: We simulated
 RCT data using diverse assumptions for the average treatment effect, a baseline
 prognostic index of risk (PI), the shape of its interaction with treatment
 (none, linear, quadratic or non-monotonic), and the magnitude of
 treatment-related harms (none or constant independent of the PI). We predicted
 absolute benefit using: models with a constant relative treatment effect;
 stratification in quarters of the PI; models including a linear interaction of
 treatment with the PI; models including an interaction of treatment with a
 restricted cubic spline (RCS) transformation of the PI; an adaptive approach
 using Akaike’s Information Criterion. We evaluated predictive performance using
 root mean squared error and measures of discrimination and calibration for
 benefit. **Results**: The linear-interaction model and the RCS-interaction
 displayed robust performance across many simulation scenarios. The RCS-model
 was optimal when quadratic or non-monotonic deviations from a constant
 treatment effect were stronger, and when sample size was larger. The adaptive
 approach required larger sample sizes. Illustrations in the GUSTO-I trial
 confirmed these findings. **Conclusion**: An interaction between baseline risk
 and treatment assignment should be considered to improve treatment effect
 predictions.
author:
  - name: Alexandros Rekkas
    affiliation: 1
    correspondingauthor: true
  - name: Peter R. Rijnbeek
    affiliation: 1
  - name: David M. Kent
    affiliation: 2
  - name: Ewout W. Steyerberg
    affiliation: 3
  - name: David van Klaveren
    affiliation: 4
address:
  - code: 1
    address: Department of Medical Informatics, Erasmus Medical Center, Rotterdam, The Netherlands
  - code: 2
    address: Predictive Analytics and Comparative Effectiveness Center, Institute for Clinical Research and Health Policy Studies, Tufts Medical Center, Boston, Massachusetts, USA
  - code: 3
    address: Department of Biomedical Data Sciences, Leiden University Medical Center, Leiden, The Netherlands
  - code: 4
    address: Department of Public Health, Erasmus Medical Center, Rotterdam, The Netherlands
numbersections: true
linenumbers: false
journal: arXiv.org
keywords:
  - treatment effect heterogeneity
  - absolute benefit
  - prediction models
output: 
  rticles::elsevier_article:
    keep_tex: true
  bookdown::word_document2:
    reference_docx: reference.docx
geometry: margin=1.0in
date: false
toc: false
font-size: 11pt
header-includes:
  - \renewcommand*\familydefault{\sfdefault}
  - \usepackage{setspace}
  - \usepackage{amsmath}
  - \doublespacing
  - \usepackage{amssymb}
  - \usepackage{bm}
  - \usepackage{caption}
  - \usepackage{booktabs}
  - \date{}
  - \newcommand\given[1][]{\:#1\vert\:}
  - \newcommand{\indep}{\perp \!\!\! \perp}
editor_options: 
  chunk_output_type: console
bibliography: references.bib
csl: jce.csl
---

@@

## Background

We observe RCT data $(Z, X, Y)$, where for each patient $Z_i= 0, 1$ is the
treatment status, $Y_i = 0, 1$ is the observed outcome and $X_i$ is a set of
covariates measured. Let $\{Y_i(z), z=0, 1\}$ denote the unobservable potential
outcomes. We observe $Y_i = Z_iY_i(1) + (1 - Z_i)Y_i(0)$. We are interested in
predicting the conditional average treatment effect (CATE), 
$$\tau(x) = E\{Y(0) - Y(1)|X=x\}$$ 
Assuming that $(Z, X, Y)$ is a random sample from the target population and that
$\big(Y(0), Y(1)\big)\indep Z|X$, as we are in the RCT setting, we can predict
CATE from
\begin{align*}
\tau(x) &= E\{Y(0)\given X=x\}-E\{Y(1)\given X=x\}\\
&=E\{Y\given X=x, Z=0\}-E\{Y\given X=x, Z=1\}
\end{align*}

Based on an estimate of baseline risk
$$
E\{Y\given X=x, Z=0\}=g\big(lp(x)\big)
$$
with $u=lp(x)=\hat{\beta}^tx$ the linear predictor and $g$ the link function,
we predict CATE from
$$
\hat{\tau}(x) = g\big(f(u, 0)\big) - g\big(f(u, 1)\big)
$$
where $f(u,z)$ describes interactions of the baseline risk linear predictor with
treatment. In the rest of the paper we use the linear predictor $lp(x)$ as the
prognostic index (PI) for the outcome $Y$.

@@

Outcomes in the treatment arm were generated using 3 base scenarios: absent
treatment effect (OR = 1), moderate treatment effect (OR = 0.8) and strong
treatment effect (OR = 0.5). We started with simulating outcomes based on true
constant relative treatment effects for the 3 base scenarios. We then simulated
linear, quadratic and non-monotonic deviations from constant treatment effects
using:
$$lp_1 = \gamma_2(PI-c)^2 + \gamma_1(PI-c) + \gamma_0, $$
where $lp_1$ is the true linear predictor in the treatment arm, so that
$P(y=1|X, Z=1) = \text{expit}(lp_1)$. Finally, we simulated scenarios where a
constant absolute harm is applied across all treated patients. In this case we
have $P(y=1|X,Z=1) = \text{expit}(lp_1) + \text{harm}$.

@@

Second, we considered a model which assumes *constant relative treatment effect*
(constant odds ratio). Hence, absolute benefit is predicted from
$\hat{\tau}(\bm{x}) = \text{expit}(PI) - \text{expit}(PI +\log(\text{OR}))$.

### 2022-04-08 submission/manuscript.tex
update manuscript and supplement

% Pandoc header
\renewcommand*\familydefault{\sfdefault}
\usepackage{setspace}
\usepackage{amsmath}
\doublespacing
\usepackage{amssymb}
\usepackage{bm}
\usepackage{caption}
\usepackage{booktabs}
\date{}
\newcommand\given[1][]{\:#1\vert\:}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}

@@

@@

Outcomes in the treatment arm were generated using 3 base scenarios:
absent treatment effect (OR = 1), moderate treatment effect (OR = 0.8)
and strong treatment effect (OR = 0.5). We started with simulating
outcomes based on true constant relative treatment effects for the 3
base scenarios. We then simulated linear, quadratic and non-monotonic
deviations from constant treatment effects using:
\[lp_1 = \gamma_2(PI-c)^2 + \gamma_1(PI-c) + \gamma_0, \] where \(lp_1\)
is the true linear predictor in the treatment arm, so that
\(P(y=1|X, t_x=1) = \text{expit}(lp_1)\). Finally, we simulated
scenarios where a constant absolute harm is applied across all treated
patients. In this case we have
\(P(y=1|X,t_x=1) = \text{expit}(lp_1) + \text{harm}\).

@@

Second, we considered a model which assumes \emph{constant relative
treatment effect} (constant odds ratio). Hence, absolute benefit is
predicted from
\(\hat{\tau}(\bm{x}) = \text{expit}(PI +\log(\text{OR}))\).

>>>>

% Pandoc header
\renewcommand*\familydefault{\sfdefault}
\usepackage{setspace}
\usepackage{amsmath}
\doublespacing
\usepackage{amssymb}
\usepackage{bm}
\usepackage{caption}
\usepackage{booktabs}
\date{}
\newcommand\given[1][]{\:#1\vert\:}
\newcommand{\indep}{\perp \!\!\! \perp}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}

@@

\hypertarget{background}{%
\subsection{Background}\label{background}}

We observe RCT data \((Z, X, Y)\), where for each patient \(Z_i= 0, 1\)
is the treatment status, \(Y_i = 0, 1\) is the observed outcome and
\(X_i\) is a set of covariates measured. Let \(\{Y_i(z), z=0, 1\}\)
denote the unobservable potential outcomes. We observe
\(Y_i = Z_iY_i(1) + (1 - Z_i)Y_i(0)\). We are interested in predicting
the conditional average treatment effect (CATE),
\[\tau(x) = E\{Y(0) - Y(1)|X=x\}\] Assuming that \((Z, X, Y)\) is a
random sample from the target population and that
\(\big(Y(0), Y(1)\big)\perp \!\!\! \perp Z|X\), as we are in the RCT
setting, we can predict CATE from \begin{align*}
\tau(x) &= E\{Y(0)\:\vert\:X=x\}-E\{Y(1)\:\vert\:X=x\}\\
&=E\{Y\:\vert\:X=x, Z=0\}-E\{Y\:\vert\:X=x, Z=1\}
\end{align*}

Based on an estimate of baseline risk \[
E\{Y\:\vert\:X=x, Z=0\}=g\big(lp(x)\big)
\] with \(u=lp(x)=\hat{\beta}^tx\) the linear predictor and \(g\) the
link function, we predict CATE from \[
\hat{\tau}(x) = g\big(f(u, 0)\big) - g\big(f(u, 1)\big)
\] where \(f(u,z)\) describes interactions of the baseline risk linear
predictor with treatment. In the rest of the paper we use the linear
predictor \(lp(x)\) as the prognostic index (PI) for the outcome \(Y\).

@@

Outcomes in the treatment arm were generated using 3 base scenarios:
absent treatment effect (OR = 1), moderate treatment effect (OR = 0.8)
and strong treatment effect (OR = 0.5). We started with simulating
outcomes based on true constant relative treatment effects for the 3
base scenarios. We then simulated linear, quadratic and non-monotonic
deviations from constant treatment effects using:
\[lp_1 = \gamma_2(PI-c)^2 + \gamma_1(PI-c) + \gamma_0, \] where \(lp_1\)
is the true linear predictor in the treatment arm, so that
\(P(y=1|X, Z=1) = \text{expit}(lp_1)\). Finally, we simulated scenarios
where a constant absolute harm is applied across all treated patients.
In this case we have
\(P(y=1|X,Z=1) = \text{expit}(lp_1) + \text{harm}\).

@@

Second, we considered a model which assumes \emph{constant relative
treatment effect} (constant odds ratio). Hence, absolute benefit is
predicted from
\(\hat{\tau}(\bm{x}) = \text{expit}(PI) - \text{expit}(PI +\log(\text{OR}))\).

### 2022-04-08 submission/supplement.rmd
update manuscript and supplement

```{r, cache=TRUE, echo=FALSE, fig.cap="Simulation scenarios that closely follow trials. In this case, we see increasing absolute benefits with increasing baseline risk.", fig.show="hold", out.width = '50%'}
include_graphics(
  c(
    here("figures/scenario_251.png"),
    here("figures/scenario_324.png")
  )
)
```

```{r, cache=TRUE, echo=FALSE, fig.cap="Simulation scenarios that closely follow trials. In this case, we see increasing absolute benefits with increasing baseline risk up to the third risk quarter. In the fourth risk quarter this trend is interrupted and benefits are diminished.", fig.show="hold", out.width = '50%'}
include_graphics(
  c(
    here("figures/scenario_406.png"),
    here("figures/scenario_422.png")
  )
)
```

>>>>

```{r, cache=TRUE, echo=FALSE, fig.cap="Simulation scenarios that closely follow trials. In this case, we see increasing absolute benefits with increasing baseline risk.", fig.show="hold", out.width = '50%'}
include_graphics(
  c(
    here("figures/scenario_251.tiff"),
    here("figures/scenario_324.tiff")
  )
)
```

```{r, cache=TRUE, echo=FALSE, fig.cap="Simulation scenarios that closely follow trials. In this case, we see increasing absolute benefits with increasing baseline risk up to the third risk quarter. In the fourth risk quarter this trend is interrupted and benefits are diminished.", fig.show="hold", out.width = '50%'}
include_graphics(
  c(
    here("figures/scenario_406.tiff"),
    here("figures/scenario_422.tiff")
  )
)
```
